Soss

### 2020-01-01 paper/paper.tex
edits

\begin{itemize}
    \item Model specification in Soss is \emph{declarative}; statements are represented not by the order entered by the user, but by the partial order given by their variable dependencies.
    \item Soss models are typically expressed in terms of \emph{inference combinators} (like ``\jl#For#'') that combine distributions in some way to arrive at a new distribution.
    \item Soss models are \emph{first-class}; models can be passed as arguments to other models, or used in place of distributions within a model.
    \item Soss models are \emph{function-like}; abstractly, a model may be considered a function from its arguments to a joint distribution. In particular, specification of ``observed variables'' is separated from model definition.
    \item Values and distributions in Soss are represented internally as \emph{abstract syntax trees}, keeping the internal representation close to that given by the user for maximum flexibility.
    \item Soss uses \emph{staged compilation} for inference primitives like \jl#rand# and \jl#logpdf#, via novel \emph{generalized generated functions} from \texttt{GeneralizedGenerated.jl}.
    \item Soss supports \emph{model transformations}, functions that take a model and return another model.
    \item Soss supports \emph{symbolic simplifcation}, making it easy to inspect or manipulate a symbolic representation of the log-density, or to use it to generate optimized code.
    \item Soss is \emph{extensible}; users can define new inference primitives or model transformations externally, and use them as if they had been included in Soss.
\end{itemize}

@@

A Soss model is \emph{function-like}, taking values for its arguments and returning a \emph{joint distribution}.

>>>>

\begin{itemize}
    \item Model specification in Soss is \emph{declarative}; statements are represented not by the order entered by the user, but by the partial order given by their variable dependencies.
    \item Soss models are typically expressed in terms of \emph{inference combinators} (like ``\jl#For#'') that combine distributions in some way to arrive at a new distribution.
    \item Soss models are \emph{first-class}; models can be passed as arguments to other models, or used in place of distribution functions like \jl#Normal# within a model.
    \item Soss models are \emph{function-like}; abstractly, a model may be considered a function from its arguments to a joint distribution. In particular, specification of ``observed variables'' is separated from model definition.
    \item Values and distributions in Soss are represented internally as \emph{abstract syntax trees}, keeping the internal representation close to that given by the user for maximum flexibility.
    \item Soss uses \emph{staged compilation} for inference primitives like \jl#rand# and \jl#logpdf#, via novel \emph{generalized generated functions} from \jl#GeneralizedGenerated.jl#.
    \item Soss supports \emph{model transformations}, functions that take a model and return another model.
    \item Soss supports \emph{symbolic simplifcation}, making it easy to inspect or manipulate a symbolic representation of the log-density, or to use it to generate optimized code.
    \item Soss is \emph{extensible}; users can define new inference primitives or model transformations externally, and use them as if they had been included in Soss.
\end{itemize}

@@

A Soss model is \emph{function-like}, taking values for its arguments and returning a \emph{joint distribution}. For example, given \jl#m# from Figure~\ref{fig:model}, we can build the joint distribution \jl#d=m(I=2, J=3)#. This supports the usual \jl#rand# and \jl#logpdf# functions, as well as some others provided by Soss described in the following sections. For example,\footnote{This and future examples are edited for space, for example rounding floating point values.}

\begin{minted}{julia}
julia> d = m(I=2, J=3); rand(d) # or rand(m(I=2, J=3))
(I = 2, J = 3, yhat = [-0.52 -1.5 0.0037; -0.81 -1.85 -0.28], σ = 2.5, λ = 1.3, σJ = 0.72, σI = 0.41, 
 β = [-0.54, -1.5, -0.023], α = [0.026, -0.26], y = [-0.45 -2.0 2.2; -3.4 -3.9 2.3])
\end{minted}
